---
title: "p8105_hw5_mk4992"
author: "Maryam Khalid"
date: "2025-11-15"
output: html_document
---

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)

```

*Problem 1: Probability of Shared Birthdays*
We investigate the probability that at least two people in a room of *n* people share a birthday. 
We assume 365 equally likely birthdays and no leap years. 
```{r}
simulate_shared_birthday <- function(n) {
birthdays <- sample(1:365, size = n, replace = TRUE)
any(duplicated(birthdays))
}

#to run a simulation with group size of n to get T/F, do
#simuluate_shared_birthday(n)

#to get probability from z runs:
#mean(replicate(z, simulate_shared_birthay(n)))

set.seed(8105)

#vector of various group sizes
group_sizes <- 2:50

#run 1000 simulations per group size
sim_results <-
  tibble(group_size = group_sizes) %>%
  mutate(
    shared_prob = map_dbl(
      group_sizes,
      ~mean(replicate(1000, simulate_shared_birthday(.x)))
    )
  )

sim_results %>%
  ggplot(aes(x = group_size, y = shared_prob)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "blue") +
  labs(
    title = "Probability of At Least One Shared Birthday",
    x = "Group Size",
    y = "Estimated Probability"
  ) +
  theme_minimal(base_size = 14)

```

The curve representing the probability of at least one shared birthday looks similar to a logistic/sigmoidal growth curve. As the group size increases, the estimated probability increases following this growth curve. This probability is near zero for single-digit sizes, rises quickly between ~15 to 30 people, amd then levels off approaching 1. The probability is 0.5 when the group size is 23 people. By n = 50, the probability is ~0.973. Shared birthdays become extremely likely way before the group size reaches 365 people. 

*Problem 2: Hypothesis Testing from a Normal Distribution*
